Handbook streamFind: backend

### 2022-07-19 handbooks/Handbook_backend_streamFind.Rmd
Large Framework update 06

---
title: "Handbook streamFind: Backend"
author: "Ricardo Cunha"
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    theme: paper
    css: "Handbook_backend_streamFind.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 9, cache = TRUE)
```
<br> <br>

@@

```{r libraries, include=FALSE}

library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)

devtools::load_all()
```

@@

@@

Trimmed example files are included in the streamFind package and are loaded below. The trimming was done with the function `trimSpectraFilesMZR` from streamFind which uses the package [mzR](https://bioconductor.org/packages/release/bioc/html/mzR.html).

```{r resource_files}
r_path <- system.file(package = "streamFind", dir = "extdata")
files <- list.files(r_path, full.names = TRUE)
```

# S4 classes

The streamFind package uses S4 class objects to define the data structure. The S4 classes, respective methods and usage are presented and exemplified through a demonstration of the basic workflow (pre-processing) for HRMS data. In short, the basic pre-processing workflow includes the following recommended steps:  (1) project creation, (2) data conversion (when needed), (3) replicate analyses assignment, (4) peak finding (or peak picking), alignment, grouping and isotopic and adduct annotation across analyses and (5) feature quality evaluation and (6) filtering.

The semantics in streamFind package is described in details in section x. Yet, a brief summary is given below for the most relevant terms:  
  * __trace__ correspond to a raw data signal (e.g., mass trace in MS data) obtained from a given analysis file;  
  * __peak__ correspond to a cluster of traces (e.g., integrated chromatographic MS peak) within an analysis file;  
  * __feature__ correspond to a group of corresponding peaks across analysis files;

## streamProject

@@

```{r newStreamProject_call, message=FALSE, results='hide'}
exS1 <- newStreamProject(files = files[1:3])
```

## msData

@@

## msAnalysis

The _msAnalysis_ class is structured as shown below. Note that the S4 method `getAnalyses` for _msData_ was used to obtain the _msAnalysis_ of the first analysis as defined by the second argument. In general, access and plotting S4 methods can also be applied directly to the _msAnalysis_ class besides _msData_ (as described in \@ref(data-access-and-visualization)) but is not part of the basic workflow and mostly used for advanced data visualization and access of low level data (e.g., traces).

@@

```{r metadata_structure}
t(metadata(exS1, analyses = 1))

#applied directly to the msAnalysis S4 class
t(metadata(exA1))
```

@@

The basic workflow consists of essential steps for processing of raw data.

`r # TODO make a scheme of the basic workflow, maybe from the PPT`

@@

```{r}
#access to peaks in msData based on built targets as used for extracting EICs
peaks(exS2_pp[c(1, 4)], mz = mz01[1], rt = rt01[1], ppm = 10, sec = 30)

@@

```{r annotation_processing, results='hide'}
#creating settings for using CAMERA
settings_an <- createSettings(
  call = "peakAnnotation",
  algorithm = "camera",
  settings = list(
    ionization = NA_character_,
    onlyIsotopes = FALSE,
    minSize = 1,
    relMinReplicates = 1,
    extraOpts = list(
      sigma = 6,
      perfwhm = 0.35,
      cor_eic_th = 0.3,
      graphMethod = "hcs",
      pval = 0.05,
      calcCiS = TRUE,
      calcIso = TRUE,
      calcCaS = TRUE,
      maxcharge = 3,
      maxiso = 5,
      ppm = 15,
      mzabs = 0.008,
      rules =  data.table::fread(system.file("rules/primary_adducts_pos.csv", package = "CAMERA"), header = TRUE),
      multiplier = 3,
      max_peaks = 500,
      intval = "maxo"
    )
  )
)

@@

```{r}
#isotopic and adduct features annotated with a given target mz and rt
t(annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE))

@@

Data filtration is crucial for proper prioritization of relevant peaks/features. In streamFind, two levels of data filtration is implemented.

@@

```{r}
#sub-setting the msData class on analyses (the 3 first)
exS2_pg[1:3, ]

#sub-setting the msData class on analys2s and features (the 3 last analyses and features annotated with a target)
target_s <- annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
exS2_pg_fl_an[4:6, target_s$id]

#sub-seeting the msData class on analys2s and features by index
exS2_pg_fl_an[1:2, 1:5]

@@

### Filter methods

The second method uses filter methods (i.e., `filterPeaks` and `filterFeatures`) and is conservative, meaning that the data is conserved but tagged as filtered (i.e., the _filtered_ column is changed to `TRUE` and the respective filter tag is added to column _filter_, as shown below).
Both filtered peaks and features can be permanently removed by running the methods `removeFilteredPeaks` or `removeFilteredFeatures`, respectively. Note that, removing features permanently does not remove filtered peaks but peaks not represented by features are filtered with "grouping" tag. 
When removing peaks, the features left without peaks representation are removed as well.
Also, the methods `peaks` and `features` have the argument _filtered_ to return filtered peaks/features. The default for _filtered_ argument is `TRUE`, meaning that filtered peaks/features are returned by default for methods `peaks` and `features`.

`r # TODO add filtered argument to peaks and features methods`

```{r}

@@

```{r}

exS2_pg <- calculateSNR(exS2_pg, targetsID = features(exS2_pg)[["id"]][1:5])

features(exS2_pg, complete = TRUE)[1:5, ]

@@

```{r exTable, echo=FALSE, eval=FALSE, include=FALSE}

# df <- data.frame(name = c("a", "b", "c", "d", "e", "f"), col1 = 1:6, col2 = 1:6)
# 
# knitr::kable(df, caption = "Example of table caption.") %>%
# kable_styling(
#   font_size = 12,
#   bootstrap_options = c("striped", "hover", "condensed", "responsive"),
#   fixed_thead = TRUE
# )

>>>>

---
title: "Handbook streamFind: backend"
author: "Ricardo Cunha"
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    theme: paper
    css: "Handbook_backend_streamFind.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", cache = TRUE) # fig.width = 9,
```
<br> <br>

@@

```{r libraries, include=FALSE, cache=FALSE}

library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(plotly)
library(streamFind)

#devtools::load_all("../")
```

@@

The streamFind package is a backend (code based) and frontend ([shiny](https://shiny.rstudio.com/) app) platform for assembly of modular workflows to support spectrometric and spectroscopic data processing. The major focus of streamFind is data processing for environmental and quality studies of the water cycle. The streamFind package aims to stimulate the use of advanced data analysis (e.g., non-target screening, statistical analysis, etc.) in routine studies, promoting standardization of data processing and structure and easing the retrospective evaluation of data. The streamFind package can be used by academics but also by technicians due to the comprehensive documentation and well categorized set of integrated functionalities (modules). This handbook focuses on describing and demonstrating the backend of the streamFind package.

@@

## Files

The streamFind package includes example files, which are located  in the folder `system.file(package = "streamFind", dir = "extdata")`. Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The files available are described in Table \@ref(tab:description-files).

```{r resource-files, cache=FALSE}
#path to files
r_path <- system.file(package = "streamFind", dir = "extdata")

#list of file full paths
files <- list.files(r_path, full.names = TRUE)
```

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  file_name = basename(files),
  type = c(rep("MS/MS", 9)),
  polarity = c(rep("positive", 9)),
  description = c(
    rep("Basic centroided MS data spiked with chemicals.", 3),
    rep("Blank centroided MS data spiked with internal standards.", 3),
    rep("Wastewater secondary effluent centroided MS data spiked with internal standards.", 3)
  )
)

knitr::kable(df_files_desc, caption = "Files included in the streamFind package.") %>%
kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE)
```

## Spiked chemicals

`r # TODO add a list of spiked chemical in the package` List the chemicals spiked in the MS data files.

# Objects

The streamFind package uses S4 class objects. The S4 classes, main methods and usage are presented and exemplified in this section. An overview of the S4 classes and their hierarchy is presented in Figure \@ref(fig:classes-overview) using MS analyses as an example. Other data types, such as from UV and Raman analyses, would lead to different classes (e.g., _uvData_ and _uvAnalysis_ and _ramanData_ and _ramanAnalysis_, respectively) but the hierarchy is similar as shown for MS data.

```{r classes-overview, echo=FALSE, fig.cap="Overview of the S4 classes and their hierarchy with emphasis on MS analyses.", out.width="120%"}
knitr::include_graphics(paste0(getwd(), "/hb_backend_figures/classes_overview.png"))
```

## Basic terminalogy

In streamFind, the terms listed below are consistently applied throughout the objects and functionalities.

* __project__ is the project title;
* __analysis__ is the name or the actual analysis file;
* __replicate__ is the name of an analysis replicate group;
* __blank__ is the name of a blank analysis replicate group ot be used for blank (blind) subtration;
* __trace__ correspond to a raw data signal (e.g., mass trace in MS data) obtained from a given analysis file;  
* __spectra__ in MS is a table with a spectrum for each cycle time (i.e., retention time);
* __EIC__ is an extracted ion chromatograms (retention time vs intensity);
* __TIC__ is a total ion chromatograms (retention time vs intensity);
* __XIC__ is a three dimensional ion chromatogram (_m/z_ vs retention time vs intensity);
* __peak__ correspond to a cluster of traces (e.g., integrated chromatographic MS peak) within an analysis file;
* __feature__ correspond to a group of corresponding peaks across analysis files;
* __MS2__ correspond to second order fragmentation data as acquired in tandem MS;
(...)

## streamProject class

@@

```{r newStreamProject-call, message=FALSE, results='hide'}
exS1 <- newStreamProject(files = files[1:3])
```

## msData class

@@

## msAnalysis class

The _msAnalysis_ class is structured as shown below. Note that the S4 method `getAnalyses` for _msData_ was used to obtain the _msAnalysis_ of the first analysis as defined by the second argument. In general, access and plotting S4 methods can also be applied directly to the _msAnalysis_ class besides _msData_ (as described in \@ref(low-level-data-access-and-plotting)) but is not part of the basic workflow and mostly used for advanced data visualization and access of low level data (e.g., traces).

@@

```{r metadata_structure}
t(streamFind::metadata(exS1, analyses = 1))

#applied directly to the msAnalysis S4 class
t(streamFind::metadata(exA1))
```

@@

The basic workflow (Figure \@ref(fig:workflow-scheme)) includes the following steps: (1) project creation, (2) data conversion (when required/applicable), (3) assignment of analysis replicate names and respective blank analysis replicates, (4) peak finding (or peak picking), (5) peak alignment and grouping into features across analyses, (6) filling of peaks missing in features, (7) annotation of isotopic and adduct features and (8) feature quality evaluation and (9) filtering. The usage of the S4 class objects within the basic workflow (pre-processing) for HRMS data is presented in this section with examples.

```{r workflow-scheme, echo=FALSE, fig.cap="Basic workflow scheme."}
knitr::include_graphics(paste0(getwd(), "/hb_backend_figures/workflow_scheme.png"))
```

@@

```{r inspecting_peaks}
#access to peaks in msData based on built targets as used for extracting EICs
peaks(exS2_pp[c(1, 4)], mz = mz01[1], rt = rt01[1], ppm = 10, sec = 30)

@@

```{r annotation_processing, results='hide', message=FALSE}
#creating settings for using CAMERA
settings_an <- createSettings(
  call = "peakAnnotation",
  algorithm = "camera",
  settings = list(
    ionization = NA_character_,
    onlyIsotopes = FALSE,
    minSize = 1,
    relMinReplicates = 1,
    extraOpts = list(
      sigma = 6,
      perfwhm = 0.35,
      cor_eic_th = 0.3,
      graphMethod = "hcs",
      pval = 0.05,
      calcCiS = TRUE,
      calcIso = TRUE,
      calcCaS = TRUE,
      maxcharge = 3,
      maxiso = 5,
      ppm = 15,
      mzabs = 0.008,
      rules =  data.table::fread(system.file("rules/primary_adducts_pos.csv", package = "CAMERA"), header = TRUE),
      multiplier = 3,
      max_peaks = 500,
      intval = "maxo"
    )
  )
)

@@

```{r annotation_results}
#isotopic and adduct features annotated with a given target mz and rt
t(streamFind::annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE))

@@

Data filtration is crucial for proper prioritization of relevant peaks/features. In streamFind, two levels of data filtration is implemented: (1) data sub-setting (permanent) and (2) filter processing steps for peaks and/or features (tagged but not removed).

@@

```{r filter_subsetting}
#sub-setting the msData class on analyses (the 3 first)
exS2_pg[1:3, ]

#sub-setting the msData class on analys2s and features (the 3 last analyses and features annotated with a target)
target_s <- streamFind::annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
exS2_pg_fl_an[4:6, target_s$id]

#sub-seeting the msData class on analyses ([i, ]) and features ([, j]) by index
exS2_pg_fl_an[1:2, 1:5]

@@

### Filter functions

The second method uses filter functions (i.e., `filterPeaks` and `filterFeatures`) and is conservative, meaning that the data is conserved but tagged as filtered (i.e., the _filtered_ column is changed to `TRUE` and the respective filter tag is added to column _filter_, as shown below). Both filtered peaks and features can be permanently removed by running the methods `removeFilteredPeaks` or `removeFilteredFeatures`, respectively. Note that, removing features permanently does not remove filtered peaks but peaks not represented by features are filtered with "grouping" tag. When removing peaks, the features left without peaks representation are removed as well. Also, the methods `peaks` and `features` have the argument _filtered_ to return filtered peaks/features. The default for the _filtered_ argument is `TRUE`, meaning that filtered peaks/features are returned by default for methods `peaks` and `features`. Simmilarly, to peakPicking, peak Grouping, etc., the filter methods use a _msSettings_ S4 class to store the filtering parameters. Also, the filtering _msSettings_ object when applied is stored in the parameters section opf the respective class (i.e., `filterPeaks` settings are stored in the _msAnalysis_ and `filterFeatures` settings are stored in the _msFeatures_).

#### Filter features

```{r filter_features}

filters <- list(
  minIntensity = 5000,
  blankThreshold = 3,
  maxReplicateIntensityDeviation = 30,
  minReplicateAbundance = 3,
  excludeIsotopes = TRUE,
  excludeAdducts = TRUE
)

settings_filFeat <- createSettings(
  call = "filterFeatures",
  algorithm = "filter",
  settings = filters
)

exS2_pg_fl_an_filFeat <- addParameters(exS2_pg_fl_an, where = "features", settings = settings_filFeat)

exS2_pg_fl_an_filFeat <- filterFeatures(exS2_pg_fl_an_filFeat)

#number of features not filtered
nrow(features(exS2_pg_fl_an, filtered = FALSE))

#number of features not filtered after applying the filters
nrow(features(exS2_pg_fl_an_filFeat, filtered = FALSE))

#remove permanently all filtered features from the msData object
exS2_pg_fl_an_filFeat_rem <- removeFilteredFeatures(exS2_pg_fl_an_filFeat, which =  "all")

exS2_pg_fl_an_filFeat_rem
```

#### Filter peaks

```{r filter_peaks}

@@

```{r quality_sn}

# exS2_pg <- calculateSNR(exS2_pg, targetsID = features(exS2_pg)[["id"]][1:5])
# 
# features(exS2_pg, complete = TRUE)[1:5, ]

@@

```{r exTable, echo=TRUE, eval=TRUE, include=TRUE}

df <- data.frame(name = c("a", "b", "c", "d", "e", "f"), col1 = 1:6, col2 = 1:6)

knitr::kable(df, caption = "Example of table caption.") %>%
kable_styling(
  font_size = 12,
  bootstrap_options = c("striped", "hover", "condensed", "responsive"),
  fixed_thead = TRUE
)
